---
params:
  dataSource: '' # set in makefile.R
  dataFile: '' # set in makefile.R
title: "Electricity Demand for Heating and Cooling in New Zealand"
subtitle: '`r paste0("Test data source: ", dataSource, " - ", dataFile)`'
author: "Ben Anderson (ben.anderson@otago.ac.nz, `@dataknut`) & Emily Jiang"
date: 'Last run at: `r Sys.time()`'
output:
  bookdown::html_document2:
    toc: TRUE
    toc_float: TRUE
    fig_caption: TRUE
    code_folding: hide
    number_sections: TRUE
  bookdown::word_document2:
    toc: TRUE
    fig_caption: TRUE
  bookdown::pdf_document2:
    toc: TRUE
    fig_caption: TRUE
    number_sections: TRUE
---

```{r knitrSetUp, include=FALSE}
knitr::opts_chunk$set(echo = TRUE) # do not echo code
knitr::opts_chunk$set(warning = TRUE)
knitr::opts_chunk$set(message = TRUE)
knitr::opts_chunk$set(fig_caption = TRUE)
knitr::opts_chunk$set(fig_height = 6) # default, make it bigger to stretch vertical axis
knitr::opts_chunk$set(fig_width = 8) # full width
knitr::opts_chunk$set(tidy = TRUE) # tidy up code in case echo = TRUE
```

# Setup

This is where we do setup stuff like attaching libraries etc.

```{r codeSetup}
# Set start time ----
startTime <- proc.time()

# Libraries ----
library(here)
projRoot <- here::here()

# Emily's libraries and functions
source(paste0(projRoot, "/R/libraries.R")) # Emily's recommended libraries
source(paste0(projRoot, "/R/functions_v1.1.R")) # Emily's functions

# Local libraries ----
library(kableExtra) # for fancy tables
library(skimr) # for data descriptors (skim)
# any extras?

# Local functions ----

```


Session info (for reference)

```{r sessionInfo}
sessionInfo()
```

# Load final data

## Heat pumps

First we load the selected pre-extracted heat pump data and run some basic tests.

```{r load.HP.data}
message("Loading", params$dataFile)

isTesting <- 0

# for testing ----
#isTesting <- 1

if(isTesting){
  params <- list() # gets set by knit
 #params$dataFile <-  paste0(myParams$dPath, "allHouseholds_hourly.csv.gz")
 params$dataFile <- "~/Data/NZ_GREENGrid/safe/gridSpy/1min/dataExtracts/Heat Pump_2015-04-01_2016-03-31_observations.csv.gz"
 params$dataSource <- "benData"
}

hpDT_orig <- data.table::fread(params$dataFile)

hpDT <- hpDT_orig

if(params$dataSource == "emilyData"){
  # drop V1 (purpose?)
  hpDT$V1 <- NULL
  # fix the date - remember it's stored as UTC
  hpDT <- hpDT[, r_dateTimeChar := r_dateTime]
  hpDT <- hpDT[, r_dateTime := lubridate::as_datetime(r_dateTime)]
  hpDT <- hpDT[, r_dateTimeLocal := lubridate::with_tz(r_dateTime, tzone = "Pacific/Auckland")] # convert to local time from UTC
}

if(params$dataSource == "benData"){
  # minute level observations, needs processing & summarising
  # calculate hourly mean
  hpDT <- hpDT[, r_dateTime := lubridate::as_datetime(r_dateTime)] # make sure date set to POSIXct
  hpDT <- hpDT[, r_dateTimeHour := lubridate::floor_date(r_dateTime, "hour")] # cuts back to hour at the start. I assume this is what Emily did
  hpDTh <- hpDT[, .(powerW = mean(powerW)), keyby = .(r_dateTime = r_dateTimeHour, circuit, linkID)] # calculate hourly mean power (temperature data is hourly means)
  # proceed as above
  hpDTh <- hpDTh[, r_dateTimeLocal := lubridate::with_tz(r_dateTime, tzone = "Pacific/Auckland")] # convert to local time from UTC
  # merge location table
  locationDT <- data.table::fread(paste0(myParams$dPath, "locationTable.csv"))
  setkey(locationDT, linkID)
  setkey(hpDTh, linkID)
  hpDT <- hpDTh[locationDT[, .(linkID, location)]]
  hpDT$id <- hpDT$linkID
}


# location is constant by id so we could just create a table and drop that columhn too but...

hpDT <- hpDT[, hms := hms::as.hms(r_dateTimeLocal)] # useful for plots and for checking we have the right timezone!

hpDT <- GREENGridData::addNZSeason(hpDT, r_dateTime = r_dateTimeLocal)
```

If any of the r_dateTimes failed to parse they will have the local time set to NA so we can see what happened (if anything) in Table \@ref(tab:hpDateError). If there are no rows in this table then all r_dateTimes parsed.

```{r hpDateError}
kableExtra::kable(head(hpDT[is.na(r_dateTime)], 10), caption = "First 10 rows of Heat Pump data where r_dateTime failed to parse (could be none at all)") %>% 
  kable_styling()
```


Table \@ref(tab:hpHead) shows the first 10 rows of slighty processed data. Note that local time ias set to NZ time while the original r_dateTime is in UTC. This is important for linking the temperature data correctly!

```{r hpHead}
kableExtra::kable(head(hpDT, 10), caption = "First 10 rows of (slightly) processed Heat Pump data") %>% 
  kable_styling()
```

Table \@ref(tab:hpExclusions) gives a list of all circuit labels seen - as a check on what we think they are measuring.

In addition, Emily's analysis excludes the following circuits:

 * `Downstairs (inc 1 Heat Pump)$2212` - this may well include a number of other appliances;
 * `Bedroom & Lounge Heat Pumps$2741` - appears to have many zero values in winter at low temperatures but so do others (see Figure \@ref(fig:powerTempByID));
 * `Theatre Heat Pump$2740` - generally low level of use (see Figure \@ref(fig:powerTempByID));

How many of them are there? Here we create two filters - one based on Emily's suggestions (3 exclusions as above) and one which allows `Bedroom & Lounge Heat Pumps$2741`.

```{r hpExclusions}

# flag the circuits we don't want
hpDT <- hpDT[, ej_includeCircuit := ifelse(circuit == "Downstairs (inc 1 Heat Pump)$2212" |
                                          circuit == "Bedroom & Lounge Heat Pumps$2741" | # why exclude this one?
                                          circuit == "Theatre Heat Pump$2740",
                                        0, # exclude 
                                        1)] # sets all others to 1

# flag the circuits we don't want
hpDT <- hpDT[, ba_includeCircuit := ifelse(circuit == "Downstairs (inc 1 Heat Pump)$2212" |
                                          circuit == "Theatre Heat Pump$2740",
                                        0, # exclude 
                                        1)] # sets all others to 1


t <- hpDT[, .(nObs = .N), keyby = .(ej_includeCircuit, ba_includeCircuit, circuit)]

kableExtra::kable(t, caption = "Counts of circuit labels for those we exclude (0) and those we include (1)") %>% 
  kable_styling()
```

Table \@ref(tab:hpHouseholds) gives summary statistics for heat pump demand for all circuits and all households (id).

```{r hpHouseholds}

t <- hpDT[, .(nObs = .N,
              meanW = mean(powerW),
              minW = min(powerW),
              maxW = max(powerW)), keyby = .(id, circuit, location)]

kableExtra::kable(t, caption = "Heat pump statistics by household") %>% 
  kable_styling()
```


Table \@ref(tab:hplocation) gives summary statistics for heat pump demand for all circuits in each location.

```{r hplocation}


t <- hpDT[, .(nHouseholds = uniqueN(id),
              nObs = .N,
              meanW = mean(powerW),
              minW = min(powerW),
              maxW = max(powerW)), keyby = .(location, season)]

kableExtra::kable(t, caption = "Heat pump statistics by household") %>% 
  kable_styling()
```

Table \@ref(tab:hpDateCounts) and Figure \@ref(fig:hpDateCounts) gives the number of households observed for all circuits in each year and season. Small numbers in any given season/year will affect the reliability of the subsequent results. On the basis of this table/plot we may wish to select specific time periods.

```{r hpDateCounts, fig.cap="Count of observed households by year and month"}


t <- hpDT[, .(nHouseholds = uniqueN(id),
              nObs = .N), keyby = .(year = as.factor(lubridate::year(r_dateTimeLocal)), month = lubridate::month(r_dateTimeLocal, label = TRUE))]

kableExtra::kable(t, caption = "Household counts by year and season") %>% 
  kable_styling()

ggplot2::ggplot(t, aes(x = year, y = month, fill = nHouseholds)) +
  geom_tile()
```

Figure \@ref(fig:simpleHistHP) shows the mean heat pump demand by time of day by season. Does this look right?

```{r simpleHistHP, fig.cap="Histogram mean hourly power demand for Heat Pumps by season and location"}

hpCap <- paste0("GREEN Grid Heat Pump circuits, n households = ", uniqueN(hpDT$id),
                    "\n Time period: ", min(lubridate::date(hpDT$r_dateTimeLocal)), 
                    " to ",
                    max(lubridate::date(hpDT$r_dateTimeLocal)))

ggplot2::ggplot(hpDT, aes(x = powerW)) +
  geom_histogram() +
  facet_grid(location ~ season) +
  labs(caption = paste0(hpCap, "\n n circuits: ", uniqueN(hpDT$circuit))
       )
```

Figure \@ref(fig:simpleHistHPno0) shows same plot but only for power > 0 W (i.e. on standby or heating) and only for the circuits BA suggests we include (all but 2 - see Table \@ref(tab:hpExclusions)). Does this look right?

```{r simpleHistHPno0, fig.cap="Histogram mean hourly power demand for Heat Pumps by season and location"}

dt <- hpDT[powerW > 0 & ba_includeCircuit == 1]
message("Check min power: ", min(dt$powerW))
ggplot2::ggplot(dt, aes(x = powerW)) +
  geom_histogram() +
  facet_grid(location ~ season) +
  labs(caption = paste0(hpCap, "\n n circuits: ", 
                        uniqueN(dt$circuit)))
```

Figure \@ref(fig:simpleHistHP40) shows same plot but only for power > 40 W and only for the circuits BA suggests we include. Does this look right?

```{r simpleHistHP40, fig.cap="Histogram mean hourly power demand for Heat Pumps by season and location"}

dt <- hpDT[powerW > 40 & ba_includeCircuit == 1]
message("Check min power: ", min(dt$powerW))
ggplot2::ggplot(dt, aes(x = powerW)) +
  geom_histogram() +
  facet_grid(location ~ season) +
  labs(caption = paste0(hpCap, "\n n circuits: ", 
                        uniqueN(dt$circuit)))
```

Figure \@ref(fig:simpleLineHP) shows the mean heat pump demand (no power value exclusions) by time of day by season for the circuits BA suggests we include. Does this look right?

Figure \@ref(fig:simpleLineHP) is important because:

 * we converted the incoming UTC dateTime to local (NZT) time -> `r_dateTimeLocal`
 * we then created a hh:mm variable from `r_dateTimeLocal` and used that to make this plot
 * so if this looks like the right shape for Heat Pump demand then we have made the correct assumptions about the incoming `r_dateTime`


```{r simpleLineHP, fig.cap="Line plot of mean power demand for Heat Pumps by season and location"}
dt <- hpDT[ba_includeCircuit == 1]

plotDT <- dt[, .(meanW = mean(powerW)), keyby = .(hms, season, location)]

ggplot2::ggplot(plotDT, aes(x = hms, y = meanW, colour = season)) +
  geom_line() +
  facet_grid(location ~ .) +
  labs(caption = paste0(hpCap, "\n n circuits: ", uniqueN(dt$circuit),
                        "\n n observations: ", nrow(dt[!is.na(powerW)]))
       )
```

Are we seeing some cooling in summer?

We can repeat this plot including only powerW values > 40 W (Figure \@ref(fig:simpleLineHP40)). This shows the level of demand _when they are on_. This plot shows some distinct signs of heat pump use in early evenings in summer producing a slightly earlier 'peak' than is the case for winter heating. It has to be said however that the plots are quite messy, no doubt due to the small numbers involved.

```{r simpleLineHP40, fig.cap="Line plot of mean power demand for Heat Pumps by season and location where power > 40W"}
dt <- hpDT[powerW > 40 & ba_includeCircuit == 1]
  
plotDT <-dt[, .(meanW = mean(powerW)), keyby = .(hms, season, location)]

ggplot2::ggplot(plotDT, aes(x = hms, y = meanW, colour = season)) +
  geom_line() +
  facet_grid(location ~ .) +
  labs(caption = paste0(hpCap, "\n Values > 40W only",
                        "\n n circuits: ", uniqueN(dt$circuit),
                        "\n n observations: ", nrow(dt[!is.na(powerW)]))
       )
```


## Temperature

### Data in 2 files

First we load this using the two seperate files.

```{r load2fTempData}
message("Loading 15876_hourly_all.txt.gz")
temp15876DT <- data.table::fread(paste0(myParams$dPath, "15876_hourly_all.txt.gz"))
message("Loading 23872_hourly_all.txt.gz")
temp23872DT <- data.table::fread(paste0(myParams$dPath, "23872_hourly_all.txt.gz"))

fixTempData <- function(dt){
  # bunch of stuff we want to do to each dataset
  # fix datetime so it matches the hp data
  dt <- dt[, r_dateTime := lubridate::dmy_hm(`Date(NZST)`, # so we know it's NZT - shame it had to have () in it!
                                                  tz = "Pacific/Auckland")] # just to be sure (if you ran this code in the UK it would set it to UTC by default)
  
  
  dt <- dt[, hms := hms::as.hms(r_dateTime)]
  
  dt <- GREENGridData::addNZSeason(dt)
  
  dt <- dt[, r_dateTimeLocal := r_dateTime] # for matching
  dt <- dt[, location := Station] # for matching
  
  return(dt)
}

temp15876DT <- fixTempData(temp15876DT)
message("Some dates failed to parse - which ones?")

kableExtra::kable(head(temp15876DT[is.na(r_dateTimeLocal), .(Station, `Date(NZST)`, r_dateTimeLocal)]), 
                  caption = "Non-parsed dates for location =  15876")


temp23872DT <- fixTempData(temp23872DT)
message("Some dates failed to parse - which ones?")

kableExtra::kable(head(temp23872DT[is.na(r_dateTimeLocal), .(Station, `Date(NZST)`, r_dateTimeLocal)]), 
                  caption = "Non-parsed dates for location =  23872")


message("Looks like the ones around the DST switches. We hate DST almost as much as we hate timezones.")


```

Convert the incoming character values to numeric as needed.

```{r checkTempData}
# combine the temperature data
tempDT <- rbind(temp15876DT, temp23872DT)

tempDT <- tempDT[, meanTemp := as.numeric(`Tmean(C)`)]

message("Note that ", nrow(tempDT[is.na(meanTemp)]), " observations have no Tmean(C) values...")

tempDT <- tempDT[, year := lubridate::year(r_dateTime)]
t <- addmargins(with(tempDT[is.na(meanTemp)], table(Station,year)))

kableExtra::kable(t, caption = "Counts of missing temperature observations by location and year (source: 2 files)") %>%
  kable_styling()

message("But it's mostly in years we don't have HP data for?")
```

Figure \@ref(fig:simplePlotTemp) shows the mean temperature by time of day by season. Does this look right? The alpha values are suggesting it's getting warmer...

As above this plot is important. If the shape looks right then we have made the correct assumptions about our incoming dateTime values.

```{r simplePlotTemp, fig.height=6, fig.cap="Line plot of mean temperature by season and location"}
plotDT <- tempDT[, .(meanT = mean(meanTemp, na.rm = TRUE)), # crucial otherwise any NA causes the calculation to be NA
                 keyby = .(hms, season, Station, year)]

tempCap <- paste0("Temperature data source: ?? NIWA ??",
                        "\n Time period: ", min(lubridate::date(tempDT$r_dateTimeLocal), na.rm = TRUE),
                        " to ",
                         max(lubridate::date(tempDT$r_dateTimeLocal), na.rm = TRUE)
                  )
  
ggplot2::ggplot(plotDT, aes(x = hms, y = meanT, colour = season, alpha = year)) +
  geom_point() +
  facet_grid(Station ~ .) +
  labs(caption = tempCap)
```

### Both datasets in 1 file

Now test the single temperature file which contains both.

```{r load1fTempData}
message("Loading alltemp_hourly.csv.gz")

tempBothDT <- data.table::fread(paste0(myParams$dPath, "alltemp_hourly.csv.gz"))

kableExtra::kable(head(tempBothDT), 
                  caption = "Head:  alltemp_hourly.csv.gz before processing")


tempBothDT <- tempBothDT[, r_dateTimeLocal := lubridate::ymd_hms(r_dateTime, # looks like NZT as the first rows it match the first few rows of temp15876DT
                                                  tz = "Pacific/Auckland")] # just to be sure (if you ran this code in the UK it would set it to UTC by default)
  
  
  tempBothDT <- tempBothDT[, hms := hms::as.hms(r_dateTimeLocal)]
  
  tempBothDT <- GREENGridData::addNZSeason(tempBothDT)
  
tempBothDT$V1 <- NULL
tempBothDT <- tempBothDT[, meanTemp := Tmean_C]


kableExtra::kable(head(tempBothDT), 
                  caption = "Head:  alltemp_hourly.csv.gz after processing")

```

Figure \@ref(fig:simplePlotTemp2) shows the mean temperature by time of day by season for the data from . Does this look right? The alpha values are suggesting it's getting warmer...

Again, this plot tells us if our dateTime assumptions are correct.

```{r simplePlotTemp2, fig.height=6, fig.cap="Line plot of mean temperature by season and location"}
# combine the temperature data

message("Note that ", nrow(tempBothDT[is.na(meanTemp)]), " observations have no Tmean(C) values...")

tempBothDT <- tempBothDT[, year := lubridate::year(r_dateTime)]
t <- addmargins(with(tempBothDT[is.na(meanTemp)], table(location,year)))

kableExtra::kable(t, caption = "Counts of missing temperature observations by location and year (source: 1 file)") %>%
  kable_styling()

message("But it's mostly in years we don't have HP data for?")

plotDT <- tempBothDT[, .(meanT = mean(meanTemp, na.rm = TRUE)), # crucial otherwise any NA causes the calculation to be NA
                 keyby = .(hms, season, location, year)]

tempCap <- paste0("Temperature data source: ?? NIWA ??",
                        "\n Time period: ", min(lubridate::date(tempBothDT$r_dateTimeLocal), na.rm = TRUE),
                        " to ",
                         max(lubridate::date(tempBothDT$r_dateTimeLocal), na.rm = TRUE)
                  )
  
ggplot2::ggplot(plotDT, aes(x = hms, y = meanT, colour = season, alpha = year)) +
  geom_point() +
  facet_grid(location ~ .) +
  labs(caption = tempCap)
```

Now we can directly compare the two sources of temperature data to see if they match. Figure \@ref(fig:matchTemp) does this by linking the two files on `r_dateTimeLocal` and `location` and then plotting the temperature values from each source.

```{r matchTemp, fig.cap="Comparison of temperature data from each source by location"}
message("N rows of 2 files merged: ", nrow(tempDT))
message("N rows of alltemp_hourly.csv.gz : ", nrow(tempBothDT))

tempDT <- tempDT[, source := "Separate files"]
tempBothDT <- tempBothDT[, source := "alltemp_hourly.csv.gz"]

# let's try just comparing dates

t2fdt <- tempDT[, .(dateTime_char_2f = `Date(NZST)`, r_dateTimeLocal, meanTemp_2files = meanTemp, location, hms_2files = hms)]
setkey(t2fdt, r_dateTimeLocal, location)

t1fdt <- tempBothDT[, .(dateTime_char_1f = r_dateTime, r_dateTimeLocal, meanTemp_1file = meanTemp, location, hms_1file = hms)]
setkey(t1fdt, r_dateTimeLocal, location)

dt <- t2fdt[t1fdt] # merge on dateTimeLocal & location

ggplot2::ggplot(dt[!is.na(r_dateTimeLocal)], aes(x = meanTemp_2files, 
                                                 y = meanTemp_1file,
                                                 colour = lubridate::hour(r_dateTimeLocal))) +
  geom_point() +
  facet_grid(. ~ location) +
  labs(caption = "Comparison of temperature data from each source by location")

# so some of them match but not all
dt <- dt[, diffTemp := meanTemp_2files - meanTemp_1file]

```

Bang on. As we can see, it makes no differnce which temperture data is used.

## Link the HP and temperature data

So it looks like the date times are OK for both the HP and temperature data and either version of the temperature data is OK too. Now let's try to link the temperature data derivd from the 2 files to the heat pump data using station and r_dateTimeLocal.

```{r link.data}
tempDT <- data.table::setkey(tempDT, location, r_dateTimeLocal) # set key to link
tempDT <- tempDT[, hour := lubridate::hour(r_dateTimeLocal)]

hpDT <- data.table::setkey(hpDT, location, r_dateTimeLocal)
hpDT <- hpDT[, hour := lubridate::hour(r_dateTimeLocal)]

# remove the NA dateTimes before we do
tempDTclean <- tempDT[!is.na(r_dateTimeLocal)]

linkedDT <- tempDTclean[hpDT] # this will add each row of tempDT to the end of each relevant row of hpDT

```

Now, because we did _not_ match on `hour` we can see if anything failed to match up in terms of time of day. There are (Table \@ref(tab:testLinkedDataHour)) but it looks like they are to do with the missing temperature data as the NAs in the temperature columns (see Table \@ref(tab:headLinkedDataHour)).

```{r testLinkedDataHour}

t <- addmargins(with(linkedDT, table(hour, i.hour, useNA = "always")))

kableExtra::kable(t, caption = "Counts of hours from each source") %>% 
  kable_styling()
```

```{r headLinkedDataHour}
t <- head(linkedDT[is.na(hour)])
kableExtra::kable(t, caption = "Rows of data where hour (from temperature data file) = NA") %>% 
  kable_styling()

```

# Test relationship between temperature and heat pump demand (BA filtered data)

So now we think we've linked the data correctly, let's do some exploratory analysis. In this section we keep the observations where power < 40 for the circuits defined in `ba_includeCircuit` (see Table \@ref(tab:hpExclusions)).

First we just plot all the hourly data for demand vs mean temperature (Figure \@ref(fig:tempDemandScatterBaFiltered)).

```{r tempDemandScatterBaFiltered, fig.cap="Plots of temperature vs demand by season and location"}

baFilteredDT <- linkedDT[ ba_includeCircuit == 1 & !is.na(meanTemp)]


t <- baFilteredDT[, .(meanW = mean(powerW),
            minW = min(powerW),
            maxW = max(powerW),
            nObs = .N,
            nCircuits = uniqueN(circuit),
            nHouseholds = uniqueN(id)), keyby = .(location)]

kableExtra::kable(t, caption = "Summary power demand statistics (all data)") %>% 
  kable_styling()

hpCap <- paste0("GREEN Grid Heat Pump circuits, n households = ", uniqueN(baFilteredDT$id),
                    "\n Time period: ", min(lubridate::date(baFilteredDT$r_dateTimeLocal)), 
                    " to ",
                    max(lubridate::date(baFilteredDT$r_dateTimeLocal)),
                "\n n circuits: ", uniqueN(baFilteredDT$circuit)
                )

tempCap <- paste0("Temperature data source: ?? NIWA ??")

combinedCaption <- paste0(hpCap, "\n", tempCap)

ggplot2::ggplot(baFilteredDT[!is.na(meanTemp)],
                aes(x = meanTemp, y = powerW/1000, colour = id)) +
  geom_point() +
  facet_grid(season ~ location) +
  labs(caption = combinedCaption,
       x = "Mean temperature",
       y = "Mean half-hourly kW")

```

That doesn't tell us a lot - although these is indeed a suggestion of their use for cooling in summer in both locations by some households. This should remind us that this analysis is based on a _very small_ sample of households. So let's try to re-create the binned box plots. In this case it is not helpful to distinguish between households but we should be aware that the patterns we see may be driven by just one or two households' behaviour.

```{r tempDemandBoxSeasonBaFiltered, fig.cap="Box plots of temperature vs demand by season and location"}
ggplot2::ggplot(linkedDT[!is.na(meanTemp)],
                         mapping = aes(x = meanTemp, y = powerW/1000)) + 
  geom_boxplot(aes(cut_width(meanTemp,
                             width = 2, # change the width to get 'wider' bins
                             boundary = 0)))  + 
  facet_grid(season ~ location) +
  labs(caption = combinedCaption,
       x = "Mean temperature",
       y = "Mean half-hourly kW")
```

Figure \@ref(fig:tempDemandBoxSeasonBaFiltered) shows binned box plots of power against demand and suggest more clearly that there is a cooling effect for households in summer - especially in 15876. Or perhaps we should more accurately say there is evidence of a summer cooling effect in the `r uniqueN(linkedDT[location == "15876"]$id)` (!) households who live in 15876. And who would base a claim on that many households?

Anyway, just for good measure lets try to re-create the U shaped plots Emily produced by ignoring season.

```{r tempDemandBoxBaFiltered, fig.cap="Box plots of temperature vs demand by season and location"}
ggplot2::ggplot(linkedDT[!is.na(meanTemp)], mapping = aes(x = meanTemp, y = powerW/1000)) + 
  geom_boxplot(aes(cut_width(meanTemp,
                             width = 2, # change the width to get 'wider' bins
                             boundary = 0)))  + 
  facet_grid(. ~ location) +
  labs(caption = combinedCaption,
       x = "Mean temperature",
       y = "Mean half-hourly kW")
```

Although we can clearly see the increased demand for heat as temperatures decline below 0, the increase in the upper temperature ranges is probably obscured by the large number of zero values. It is arguable whether leaving the zeros out is valid since they could be interpreted as reflecting a household's lack of need for cooling. Were we to only include values > 40 W (to exclude 0 and standby power values) we would be analysing (and subsequently modelling) only those households who actually responded to the temperature change by demanding cooling - which will not be the case for all households in the population.


# Test relationship between temperature and heat pump demand (EJ filtered data)

Notwithstanding the previous argument re data exclusion, in this section we test Emily's suggestion to only keep heat pump data where power > 40 for the circuits defined in `ēj_includeCircuit` (see Table \@ref(tab:hpExclusions)).

First we just plot all the hourly data for demand vs mean temperature (Figure \@ref(fig:tempDemandScatterEjFiltered)).

```{r tempDemandScatterEjFiltered, fig.cap="Plots of temperature vs demand by season and location"}
ejFilteredDT <- linkedDT[powerW > 40 & ej_includeCircuit == 1 & !is.na(meanTemp)]

t <- ejFilteredDT[, .(meanW = mean(powerW),
            minW = min(powerW),
            maxW = max(powerW),
            nObs = .N,
            nCircuits = uniqueN(circuit),
            nHouseholds = uniqueN(id)), keyby = .(location)]

kableExtra::kable(t, caption = "Summary power demand staistics (filtered)") %>% 
  kable_styling()

hpCap <- paste0("GREEN Grid Heat Pump circuits, n households = ", uniqueN(ejFilteredDT$id),
                    "\n Time period: ", min(lubridate::date(ejFilteredDT$r_dateTimeLocal)), 
                    " to ",
                    max(lubridate::date(ejFilteredDT$r_dateTimeLocal)),
                "\n n circuits: ", uniqueN(ejFilteredDT$circuit)
                )

tempCap <- paste0("Temperature data source: ?? NIWA ??")

combinedCaption <- paste0(hpCap, "\n", tempCap)

ggplot2::ggplot(ejFilteredDT, aes(x = meanTemp, y = powerW/1000, colour = id)) +
  geom_point() +
  facet_grid(season ~ location) +
  labs(caption = combinedCaption,
       x = "Mean temperature",
       y = "Mean half-hourly kW")

```

Again, that doesn't tell us a lot - although these is indeed a suggestion of their use for cooling in summer in both locations. So let's re-try to create the binned box plots.

```{r tempDemandBoxSeasonEjFiltered, fig.cap="Box plots of temperature vs demand by season and location"}
ggplot2::ggplot(ejFilteredDT, mapping = aes(x = meanTemp, y = powerW/1000)) + 
  geom_boxplot(aes(cut_width(meanTemp,
                             width = 2, # change the width to get 'wider' bins
                             boundary = 0)))  + 
  facet_grid(season ~ location) +
  labs(caption = combinedCaption,
       x = "Mean temperature",
       y = "Mean half-hourly kW")
```

Figure \@ref(fig:tempDemandBoxSeasonEjFiltered) shows binned box plots of power against demand. As we would expect, the plot suggests more clearly that there is a cooling effect for households in summer when we remove 0 and power < 40 W values.

Anyway, lets again try to re-create the U shaped plots Emily produced by ignoring season.

```{r tempDemandBoxEjFiltered, fig.cap="Box plots of temperature vs demand by season and location"}
ggplot2::ggplot(ejFilteredDT, mapping = aes(x = meanTemp, y = powerW/1000)) + 
  geom_boxplot(aes(cut_width(meanTemp,
                             width = 2, # change the width to get 'wider' bins
                             boundary = 0)))  + 
  facet_grid(. ~ location) +
  labs(caption = combinedCaption,
       x = "Mean temperature",
       y = "Mean half-hourly kW")
```

The increase in demand for power as the temperature rises is now clearer.

> But this is fairly obvious since we are just looking at the households who responded to the temperature changes...

# Modelling

Here I am trying to attempt making a quadratic model using the _whole_ temperature range.
```{r ejFilteredModelPolyLMTotal}
polyTemp <- poly(ejFilteredDT$meanTemp, 2)
lmTemp <- lm(ejFilteredDT$powerW~polyTemp) # quadratic - TOTAL DATA\
summary(lmTemp)

```
But I realise I I can't use each point to model the overall trend - need to use mean or median powerW at each temperature.

So here I've made a table with the temp, mean power for that temp, and median power for that temp which can be used to model the overall trend.
``` {r modelDT}
createMeanMedianPower <- function(dt, temp, width){
  newDT <- dt[temp - 0.5*width <= dt$meanTemp & dt$meanTemp < temp + 0.5*width]
  meanTempMeanPower <- mean(newDT$powerW, na.rm = T)
  meanTempMedianPower <- median(newDT$powerW, na.rm = T)
  return(data.table("temp" = temp,
             "meanPowerW" = meanTempMeanPower,
             "medianPowerW" = meanTempMedianPower))
}

modelDT <- rbindlist(lapply(seq(min(ejFilteredDT$meanTemp), max(ejFilteredDT$meanTemp), 1), function(l){
      return(createMeanMedianPower(ejFilteredDT, l, 1))
    }))
```

Here I am making the linear model part for the mean.
``` {r linearModelMean}
  polyTemp <- poly(modelDT$temp, 2)
  lmMeanPowerW <- lm(modelDT$meanPowerW~polyTemp)
  summary(lmMeanPowerW)

```

Here I am making the linear model part for the median.
``` {r linearModelMedian}
  lmMedianPowerW <- lm(modelDT$medianPowerW~polyTemp)
  summary(lmMedianPowerW)

```

Here I am making the fit of the models (both mean and median).
``` {R fitted}
modelDT <- cbind(modelDT, meanPowerFit = fitted(lmMeanPowerW), medianPowerFit = fitted(lmMedianPowerW))
modelDT

```
tbc

> Needs to include the 0 values (see discussion above)?

> If the regression model assumes independence of observations then this is violated if we have multiple observations for each house (which we do whether we just use the ‘best’ year or not…). The short-term fix would be to use robust standard errors in model reporting or maybe a fixed-effects model (?) but that might be well beyond scope of the report!

# Runtime

```{r check.runtime}
t <- proc.time() - startTime

elapsed <- t[[3]]
```

Analysis completed in `r elapsed` seconds ( `r round(elapsed/60,2)` minutes) using [knitr](https://cran.r-project.org/package=knitr) in [RStudio](http://www.rstudio.com) with `r R.version.string` running on `r R.version$platform`.

# R packages used

> list packages used & cite
 
# Annexes

## Detailed statistics

Figure \@ref(fig:powerTempByID) shows the relationship between mean temperature and mean power for all hourly observations for all circuits colour coded by season. Smoothed fit lines have been added using `mgcv:gam()` curves due to the size of the dataset (see `?geom_smooth` and [@ggplot]).

```{r powerTempByID, fig.cap="Pairs plot of temperature vs heat pump power demand for all households"}
ggplot2::ggplot(linkedDT, aes(x = meanTemp, y = powerW, colour = season)) +
  geom_point(alpha = 0.5) +
  geom_smooth() +
  facet_wrap(. ~ circuit)

```

## Data codebooks

### Original power data (before processing)

Depending on input files choice this will vary.

 * data source: `r params$dataSource`
 * data file: ``r params$dataFile`

```{r skimHPorig}
head(hpDT_orig)

skimr::skim(hpDT_orig)
```


### Power data (after processing)

```{r skimHP}
skimr::skim(hpDT)
```

### Original temperature file(s)

File 1:

```{r skimT1}
skimr::skim(temp15876DT)
```

File 2:

```{r skimT2}
skimr::skim(temp23872DT)
```

Combined file:

```{r skimTBoth}
skimr::skim(tempBothDT)
```

# References


