---
title: "Electricity Demand for Heating and Cooling in New Zealand"
subtitle: "Test outputs"
author: "Ben Anderson (b.anderson@soton.ac.uk, `@dataknut`) & Emily Jiang
date: 'Last run at: `r Sys.time()`'
output:
  bookdown::html_document2:
    fig_caption: yes
    code_folding: hide
    number_sections: yes
  bookdown::pdf_document2:
    fig_caption: yes
    number_sections: yes
  bookdown::word_document2:
    fig_caption: yes
    number_sections: yes
---
```{r knitrSetUp, include=FALSE}
knitr::opts_chunk$set(echo = FALSE) # do not echo code
knitr::opts_chunk$set(warning = FALSE)
knitr::opts_chunk$set(message = FALSE)
knitr::opts_chunk$set(fig_caption = TRUE)
knitr::opts_chunk$set(fig_height = 6) # default, make it bigger to stretch vertical axis
knitr::opts_chunk$set(fig_width = 8) # full width
knitr::opts_chunk$set(tidy = TRUE) # tidy up code in case echo = TRUE
```

# Setup

```{r codeSetup, include=FALSE}
# Set start time ----
startTime <- proc.time()

# What are we running on?
sysInfo <- Sys.info()
myParams <- list()
myParams$sysName <- sysInfo[[1]]
myParams$nodeName <- sysInfo[[4]]
myParams$userName <- sysInfo[[7]]

if(myParams$userName == "ben"){ # BA laptop
  myParams$dPath <- path.expand("/Volumes/hum-csafe/Research Projects/GREEN Grid/Self_contained_Projects/2018_coolingEmily/archive/")
}else{
  myParams$dPath <- "Z:/Research Projects/GREEN Grid/Self_contained_Projects/2018_coolingEmily/archive/" # <- windoze
}
  
# Functions ----

# Give feedback ----
message("Running on ", myParams$sysName)

# Emily's libraries and functions
source("../R/libraries.R") # Emily's recommended libraries
source("../R/functions.R") # Emily's functions

# Local libraries ----
library(kableExtra)
# any extras?

```
# Load final data

## Heat pumps
First we load Emily's processed heat pump data and run some basic tests.

```{r load.HP.data}
hpDT <- data.table::fread(paste0(myParams$dPath, "allHouseholds_hourly.csv"))

# fix the date - remember it's UTC
hpDT <- hpDT[, r_dateTime := lubridate::as_datetime(r_dateTime)]
hpDT <- hpDT[, r_dateTimeLocal := lubridate::with_tz(r_dateTime, tzone = "Pacific/Auckland")] # convert to local time from UTC

hpDT <- hpDT[, hms := hms::as.hms(r_dateTimeLocal)]

hpDT <- GREENGridData::addNZSeason(hpDT, r_dateTime = r_dateTimeLocal)

kableExtra::kable(head(hpDT, 10), caption = "First 10 rows of (slightly) processed Heat Pump data") %>% 
  kable_styling()

kableExtra::kable(summary(hpDT), caption = "Summary of Heat Pump data") %>% 
  kable_styling()

t <- table(hpDT$circuit)

kableExtra::kable(t, caption = "Counts of circuit labels") %>% 
  kable_styling()
```

Figure \@ref(fig:simplePlot_HP) shows the mean heat pump demand by time of day by season. Does this look right?

```{r simplePlot_HP, fig.cap="Line plot of mean power demand for Heat Pumps"}
plotDT <- hpDT[, .(meanW = mean(powerW)), keyby = .(hms, season)]

ggplot2::ggplot(plotDT, aes(x = hms, y = meanW, colour = season)) +
  geom_line()
```

Are we seeing some cooling in summer?

## Temperature

First we load this using the two seperate files.



# Runtime

```{r check runtime}
t <- proc.time() - startTime

elapsed <- t[[3]]
```

Analysis completed in `r elapsed` seconds ( `r round(elapsed/60,2)` minutes) using [knitr](https://cran.r-project.org/package=knitr) in [RStudio](http://www.rstudio.com) with `r R.version.string` running on `r R.version$platform`.

# R environment

> list packages used & cite
 
```{r sessionInfo}
sessionInfo()
```

